The role of advection versus diffusion in intraventricular transport

### 2025-02-10 paper/main.tex
minor updates
add shrunk cross-sectional areas information
edits after Nathalie's revision

% Title
\title{The role of advection versus diffusion in intraventricular transport}
\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[FIX ME]{Inyoung Jeong} % OK?
\author[1,3,+]{Marie E.~Rognes} % OK order?
\author[2,+, *]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing,
Simula Research Laboratory, Oslo, Norway}
\affil[2]{\fixme{@Nathalie is this correct?: Department of Clinical and Molecular Medicine,
Norwegian University of Science and Technology, Trondheim, Norway\\
Same affiliation for Inyoung?}}
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}
\href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\affil[+]{these authors contributed equally to this work}
% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={zebrafish-cilia-csf},
  pdfauthor={H.~Herlyng~\emph{et al.}}
}
\fi

\begin{abstract}
    This paper presents ...
    %Flow of cerebrospinal fluid in zebrafish brain ventricles is studied in this work. Motivated by experimental observations, we consider motile cilia and cardiac beating as the driving flow mechanisms.
    %A mathematical model of the flow is developed and coupled with a transport model to study the transport of molecules inside the brain ventricles. 
    %The Stokes equations are used to model the flow, with the motile cilia acting as a force on the fluid through a stress boundary condition that allows slip at the wall. 
    %Transport is modeled with an advection-diffusion equation coupled to the velocity field obtained from solving the Stokes equations.
    %The flow and transport models are discretized with finite element methods. For the Stokes equations we employ a non-conforming discretization based on elements of the Brezzi-Douglas-Marini family. 
    %The advection-diffusion equation is discretized with discontinuous Galerkin elements. 
    %We implement and solve the discretized equations numerically to simulate flow and transport in a zebrafish brain ventricles geometry obtained with medical imaging.
\end{abstract}
\begin{document}
\maketitle
% Introduction:
% 1. What are cilia and why are they important?
% 2. Previous studies of cilia flow
% 3. What problem do we want to solve
% 4. The model and work done in this paper
\section*{Introduction}
Brain development and function depend on numerous regulatory processes,
including interactions with cerebrospinal fluid (CSF).
CSF is a clear liquid produced by the choroid plexus, which is a secretory
epithelium located within the brain ventricles.
Following production, CSF flows through the ventricular system and
subarachnoid space, where it regulates brain development and
homeostasis by transporting growth factors, nutrients, neurotransmitters,
waste products and other molecules and metabolites.

To achieve its function, CSF needs to circulate throughout the brain.
In fact, CSF disturbances can univocally lead to brain disorders,
such as hydrocephalus where accumulation of CSF requires emergency surgical treatment.
CSF movement occurs through multiple physiological processes, including the action of
motile cilia on the ventricular wall,
CSF secretion by the choroid plexus and pressure gradients generated by the
cardiac cycle, respiration, and bodily movement.
Motile cilia are microscopic hair-like structures that beat in a metachronal
wave pattern to generate flow pattern. Due to their microscopic size, ranging from 6 to 15 microns,
they contribute to fluid flow primarily at the surface of the brain ventricles.
In contrast, pressure gradients generate fluid flow further from the ventricular
walls and along the perivascular space surrounding the major blood vessels.

To date, it remains poorly understood how cilia contribute to the overall fluid
flow and solute transport in the brain.
These questions have been especially challenging to tackle due to technical limitations.
Notably, cilia are very small and beat at high frequencies ranging from 10 to 30 Hz,
while other fluid dynamic processes in the brain are slower and span larger volumes.
Most studies until now have analyzed one process at a time, especially in mammals.
For instance, multiple studies have convincingly shown that cilia generate complex
flow patterns at the surface of the brain ventricles.
However, since all these studies have been performed on brain explants, they failed
to identify the role of pressure gradients associated with respiration and heart pulsations in CSF transport.
Similarly, while MRI technology can measure CSF movement in humans, it cannot detect
the fast movements of cilia or understand their impact in the large human brain.
Consequently, we lack a comprehensive understanding of how cilia contribute to CSF
movement and solute transport, and whether cilia dysfunction will alter CSF distribution
and lead to brain disorders.
This is particularly important since ventricular enlargement or hydrocephalus is
commonly observed in animal models with motile cilia deficiency, potentially affecting
brain activity and behavior.

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Cilia have been studied previously,
with numerous studies on the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
coral reefs~\cite{Pacherres2022CiliaryProduction}, and the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These studies have provided valuable insights, showing that cilia lend themselves to simulation studies.

In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in CSF flow and solute transport.
First, we computed flow patterns and intraventricular transport \emph{in silico}. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF movement velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages, while still being well-conserved with mammals.
Next, we validated our computation model with experimental data by monitoring
the dynamics of a photoconvertible fluorescent protein expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia
collaborate with other factors to distribute molecules and
particles within the brain ventricles.

@@

The morphology and geometry of the brain ventricles vary across zebrafish
individuals, and under physiological and pathological conditions. 
Motivated by an interest in how ventricular geometry impacts solute distribution,
we also consider four variations in the geometry \fixme{(percentage reduction in cross-sectional areas):
by shrinking the fore-mid brain connection (66 \%), 
the middle ventricle (33 \%), the mid-hind brain connection (43 \%), and all three simultaneously.} 
We used Blender~\cite{Community2018BlenderPackage} to modify the brain ventricles geometry.

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:fig1}\fixme{a--c}).
In addition, cardiac pulsations induces pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4} \ \mathrm{Pa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

@@

\subsection*{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development,
we study the transport of molecules resembling the Dendra2 Fluorescent Protein (Dendra2)
used in this study, the Starmaker+Green Fluorescent Protein (STM+GFP) reported in
Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}, and Exosomes.
To estimate the diffusion coefficients of these, when and if not
available in the literature, we use the Stokes--Einstein relation:
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical
molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\eqref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant,
$T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the
solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\eqref{eq:D_stokes_einstein}
with $R=200$ nm. This radius covers the slowest diffusing exosomes,
since the smaller exosomes would have diffusion coefficients similar to the other molecules.
For STM+GFP, we estimated the diffusion coefficient by extrapolating
the experimentally observed diffusion coefficient of
GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\eqref{eq:D_stokes_einstein},
$D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations
are $D_1$ (Exosomes), $D_2$ (STM+GFP) and $D_3$ (Dendra2)~(\Cref{tab:diff_coeff_table}).

@@

%%------PHYSICAL EXPERIMENTS------%%
\subsection*{Zebrafish maintenance and strains in photoconversion experiments}
The animal facilities for zebrafish~(\emph{danio rerio}) are approved by the
Norwegian Food Safety Authority (NFSA, Mattilsynet).
The zebrafish were maintained in accordance with the guidelines set by the
NFSA and the European Communities Council Directive.

@@

\subsection*{Genotyping of adult zebrafish}
Genomic DNA (gDNA) was isolated from clipped fins of anesthetized adult
fish using 100 \textmu L of PCR lysis buffer (containing 1M tris pH 7--9, 0.5 M EDTA,
tritonX-100, and Proteinase K 0.1 mg/ml) overnight at 50°C.
To stop the lysis reaction, the samples were heated to 95°C for 10 minutes
and then centrifuged at 13000 rpm for 2 minutes.
The supernatant containing gDNA was utilized for KASP assays-based analysis.
The gDNAs were diluted (1:2) with water, and 3 \textmu L of the diluted gDNA was
used to perform the KASP assay following the guidelines of the manufacturer (LGC Biosearch Technologies™).
For each sample well, the master mix contained: 5 \textmu L of master mix; 0.14
\textmu L of assay mix; 1.86 \textmu L of RO water.

\subsection*{Generation of \emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}} line}
\fixme{@Nathalie: This needs a more understandable title for us non-experts, could we change to
e.g. "Generation of transgenic line"?.}
The open reading frame (ORF) of the neuropeptide y (npy) signal peptide
(\emph{npy}: ENSDARG00000036222, Q1LW93) was fused with the N-terminal
of zebrafish codon optimized Dendra2 DNA sequence.
The DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and
3’ ends (GenScript Biotech), and inserted into the
pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish}
through restriction enzyme cloning (GenScript Biotech).

To generate the transgenic line, the 5xuas:Signal-Dendra2 plasmid DNA was used.
A volume of 2 nl of a mixture of the plasmid DNA (60 pg) and tol2 mRNA (10 pg)
was microinjected into one-cell stage embryos,
as described in Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}. The injected
embryos were raised to adulthood (F0).
Germline-transmitted founder zebrafish were identified by breeding with
multiple Gal4 transgenic lines. Stable F1 embryos expressing the Gal4-driven
Dendra2 signals were screened and raised to adult zebrafish.

@@

We seek to model the compartimentalized and pulsatile CSF flow observed \emph{in vivo}~(\Cref{fig:fig1}\fixme{b--c}).
To characterize CSF flow patterns predicted by
computational fluid dynamics within the brain ventricles, we begin by
considering the simulated CSF velocity and pressure when only considering
cilia motion, disregarding cardiac pressure pulsations. The
bidirectional traction induced by the motile cilia
inherently leads to a partial compartmentalization of the ventricular
system, with large-scale vortex structures in the anterior and middle
ventricles~(\Cref{fig:fig1}\fixme{e}). The highest velocities ($26.5$ \textmu m/s)
occur in the vicinity of the dorsal cilia in the middle ventricle.

@@

Next we consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}\fixme{e},
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles are dominated by
the cilia contribution, while the cardiac pressure pulsations dominate
within the posterior ventricle. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, as a
result of both the traction induced by the cilia and the pulsatile
pressure-driven flow being aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale, yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed \emph{in vivo}.

>>>>

% Title
\title{The role of advection versus diffusion in intraventricular transport}
\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong} % OK?
\author[1,3,+]{Marie E.~Rognes} % OK order?
\author[2,+,*]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing,
Simula Research Laboratory, Oslo, Norway}
\affil[2]{Department of Clinical and Molecular Medicine,
Norwegian University of Science and Technology, Trondheim, Norway}
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}
\href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\affil[+]{these authors contributed equally to this work}
% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={zebrafish-cilia-csf},
  pdfauthor={H.~Herlyng~\emph{et al.}}
}
\fi

\begin{abstract}
    This paper presents ...
    %Flow of cerebrospinal fluid in zebrafish brain ventricles is studied in this work. Motivated by experimental observations, we consider motile cilia and cardiac beating as the driving flow mechanisms.
    %A mathematical model of the flow is developed and coupled with a transport model to study the transport of molecules inside the brain ventricles. 
    %The Stokes equations are used to model the flow, with the motile cilia acting as a force on the fluid through a stress boundary condition that allows slip at the wall. 
    %Transport is modeled with an advection-diffusion equation coupled to the velocity field obtained from solving the Stokes equations.
    %The flow and transport models are discretized with finite element methods. For the Stokes equations we employ a non-conforming discretization based on elements of the Brezzi-Douglas-Marini family. 
    %The advection-diffusion equation is discretized with discontinuous Galerkin elements. 
    %We implement and solve the discretized equations numerically to simulate flow and transport in a zebrafish brain ventricles geometry obtained with medical imaging.
\end{abstract}
\begin{document}
\maketitle
% Introduction:
% 1. What are cilia and why are they important?
% 2. Previous studies of cilia flow
% 3. What problem do we want to solve
% 4. The model and work done in this paper
\section*{Introduction}
Brain development and function depend on numerous regulatory processes,
including interactions with cerebrospinal fluid
(CSF). CSF is a clear liquid produced by the choroid plexus
that flows through the ventricular system and subarachnoid space.
By delivering growth factors, nutrients, neurotransmitters and metabolites
and removing waste products, CSF regulates brain development and homeostasis.
Disturbances of CSF circulation can lead to brain disorders, such as
hydrocephalus where accumulation of CSF often requires emergency surgical treatment.

CSF movement occurs through multiple physiological processes,
including the action of motile cilia on the ventricular walls,
CSF secretion by the choroid plexus, and pressure gradients generated
by the cardiac cycle, respiration, and bodily movement. Motile cilia
are microscopic hair-like structures that beat in a metachronal wave
pattern to mediate fluid flow. Due to their microscopic size, ranging
from 6 to 15 microns, they contribute to flow primarily at the surface
of the brain ventricles. In contrast, pressure gradients induce fluid
flow \fixme{further from the ventricular walls (bulk flow)} and along the
perivascular space surrounding the major blood vessels.

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport in the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and heart pulsations in CSF
transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
movement and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
\fixme{potentially affecting brain activity and behavior.}

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
coral reefs~\cite{Pacherres2022CiliaryProduction}, and the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia \fixme{pump fluids}
and may contribute to tissue physiology.

In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in CSF flow and solute transport.
First, we computed flow patterns and intraventricular transport \emph{in silico}. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF movement velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages, while still being well-conserved with mammals.
Next, we validated our computation model with experimental data by monitoring
the dynamics of a photoconvertible fluorescent protein Dendra2 expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia
collaborate with other factors to distribute molecules and
particles within the brain ventricles.

@@

The morphology and geometry of the brain ventricles vary across zebrafish
individuals, and under physiological and pathological conditions. 
Motivated by an interest in how ventricular geometry impacts solute distribution,
we also consider four variations in the geometry \fixme{(percentage reduction in cross-sectional areas):
by shrinking the fore-mid brain connection (66 \%), 
the middle ventricle (33 \%), the mid-hind brain connection (43 \%), and all three simultaneously.} 
We used Blender~\cite{Community2018BlenderPackage} to modify the brain ventricles geometry.
% Cross-sectional areas of ducts normal to flow direction
% Modified region       Original mesh        Deformed mesh
% Fore                  3.36e-3 mm^2         1.15e-3 mm^2
% Mid                   6.11e-3 mm^2         4.08e-3 mm^2
% Hind                  2.91e-3 mm^2         1.67e-3 mm^2

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:fig1}\fixme{a--c}).
In addition, cardiac pulsations induce pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4} \ \mathrm{Pa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

@@

\subsection*{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development,
we study the transport of molecules resembling the Dendra2 Fluorescent Protein (Dendra2)
used in this study, the Starmaker+Green Fluorescent Protein (STM+GFP) reported in
Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}, and exosomes.
To estimate the diffusion coefficients of these, when not
available in the literature, we use the Stokes--Einstein relation:
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical
molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\eqref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant,
$T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the
solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\eqref{eq:D_stokes_einstein}
with $R=200$ nm. This radius covers the slowest diffusing exosomes,
since the smaller exosomes would have diffusion coefficients similar to the other molecules.
For STM+GFP, we estimated the diffusion coefficient by extrapolating
the experimentally observed diffusion coefficient of
GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\eqref{eq:D_stokes_einstein},
$D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations
are $D_1$ (exosomes), $D_2$ (STM+GFP) and $D_3$ (Dendra2)~(\Cref{tab:diff_coeff_table}).

@@

%%------PHYSICAL EXPERIMENTS------%%
\subsection*{Zebrafish maintenance, genotyping and strains in photoconversion experiments}
The animal facilities for zebrafish~(\emph{danio rerio}) are approved by the
Norwegian Food Safety Authority (NFSA, Mattilsynet).
The zebrafish were maintained in accordance with the guidelines set by the
NFSA and the European Communities Council Directive.

@@

For genotyping, genomic DNA (gDNA) was isolated from clipped fins of anesthetized adult
fish using 100 \textmu L of PCR lysis buffer (containing 1M tris pH 7--9, 0.5 M EDTA,
tritonX-100, and Proteinase K 0.1 mg/ml) overnight at 50°C.
To stop the lysis reaction, the samples were heated to 95°C for 10 minutes
and then centrifuged at 13000 rpm for 2 minutes.
The supernatant containing gDNA was utilized for KASP assays-based analysis.
The gDNAs were diluted (1:2) with water, and 3 \textmu L of the diluted gDNA was
used to perform the KASP assay following the guidelines of the manufacturer (LGC Biosearch Technologies™).
For each sample well, the master mix contained: 5 \textmu L of master mix; 0.14
\textmu L of assay mix; 1.86 \textmu L of RO water.

\subsection*{Generation of transgenic line to study CSF \fixme{and transport} dynamics} %\emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}}
To generate a transgenic line expressing the photoconvertible protein Dendra2 in
the brain ventricles, we fused the open reading frame (ORF)
of the neuropeptide y (npy) signal peptide
(\emph{npy}: ENSDARG00000036222, Q1LW93) to the N-terminal of
zebrafish codon optimized Dendra2 DNA sequence.
The DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and
3’ ends (GenScript Biotech) and inserted into the
pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish}
through restriction enzyme cloning (GenScript Biotech)
to generate the 5xuas:Signal-Dendra2 plasmid DNA..

A volume of 2 nl of a mixture of the plasmid DNA (60 pg) and tol2 mRNA (10 pg)
was microinjected into one-cell stage embryos to generate the transgenic line,
as described in Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}. The injected
embryos were raised to adulthood (F0).
Germline-transmitted founder zebrafish were identified by breeding with
multiple Gal4 transgenic lines. Stable F1 embryos expressing the Gal4-driven
Dendra2 signals were screened and raised to adult zebrafish.

@@

We model the compartmentalized and pulsatile CSF flow observed \emph{in vivo}~(\Cref{fig:fig1}\fixme{b--c})
by the Stokes equations, and solve the equations numerically using a finite element method.
To characterize CSF flow patterns predicted by
computational fluid dynamics within the brain ventricles, we first
consider the simulated CSF velocity and pressure when only modeling
cilia motion, disregarding cardiac pressure pulsations. The
bidirectional traction induced by the motile cilia
inherently leads to a partial compartmentalization of the ventricular
system, with large-scale vortex structures in the anterior and middle
ventricles~(\Cref{fig:fig1}\fixme{e}). The highest velocities ($26.5$ \textmu m/s)
occur in the vicinity of the dorsal cilia in the middle ventricle.

@@

Next we consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}\fixme{e},
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles is majorly driven
by cilia, while the cardiac pressure pulsations dominate
within the posterior ventricle and in the ducts connecting the ventricular
compartments. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, since 
both the tangential traction induced by the cilia and the pulsatile
pressure-driven flow are aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed \emph{in vivo}~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Ventricular solute transport is advection-dominated for larger molecules}
We next explore the impact of these CSF flow patterns and magnitudes
on solute transport within the brain ventricles.
To this end, we simulate the distribution and evolution
of different solutes subjected to photoconversion within a region of
interest in the dorsal region of the middle ventricle (ROI 1)
(\Cref{fig:fig2}a). We consider the spread of
a solute concentration with a given molecular diffusion coefficient,
corresponding to exosomes ($D_1 = 1.63 \times 10^{-12}$ m$^2$/s),
a circa 90 kDa protein STM-GFP ($D_2 = 5.75 \times 10^{-11}$ m$^2$/s),
and the 26 kDa photoconvertible protein  Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).
As expected, transport occurs more
rapidly with increasing diffusion coefficient. Interestingly, the
distribution patterns differ between the solutes.
For the smallest diffusion coefficient, representing
exosomes, the advective transport by the CSF flow dominates diffusion
(\Cref{fig:fig2}b). The vortex structures of the velocity field
carry over into the concentration field, indicating that transport
along the streamlines is more rapid than the diffusion across the
streamlines. The balance between diffusion and advection shifts with
increasing diffusion coefficients: the CSF flow structures become more
obfuscated, and the solute spreads more uniformly throughout the
ventricular geometry (\Cref{fig:fig2}c, d). Notably, the
smaller solutes with higher diffusion coefficients spread more quickly
to regions distal to the photoconversion site.
\begin{figure}%[H]
    \centering
    \includegraphics[width=0.95\textwidth]{graphics/figure2_simulation_results_different_coeffs.png}
    \caption{Simulated distribution and evolution after photoconversion of
    exosomes ($D_1$), STM+GFP ($D_2$) and Dendra2 ($D_3$).
    \textbf{a.} Geometry-centered clip of the ventricles mesh showing the locations of the ROIs.
    \textbf{b--d.} Slice of the concentration field simulated
    with diffusion coefficients $D_1$, $D_2$ and $D_3$ for eight time instants covering the
    time range from 45 seconds to 790 seconds, corresponding to 100 to 1750 cardiac cycles.
    Slices are taken in the $xz$-plane ($y=0.140$ mm, the center of the geometry).
    The colorbar is scaled so the maximum is the maximum concentration of the Dendra2 
    concentration field slice at the last time instant.
    \textbf{e--g.} Mean concentration $\cbar$ in each region of interest (ROI)
    as function of time for diffusion coefficient $D_1$, $D_2$ and $D_3$, respectively. 
    \textbf{h.} The time $\hat{t}$ before the mean concentration $\cbar$ exceeds a
    threshold value of $0.25$ (ROIs 2-4) or $0.10$ (ROIs 5, 6) in each simulation setup.
    Note that there is no bar for $D_1$ in ROI 6, because the threshold value was never exceeded.}
    \label{fig:fig2}
\end{figure}
